## Demo: Basic statistical analyses


Let us start by illustrating how to peform simple statistical data analyses using different resources. Here, we will use data from three studies that are available in our Opal demo repository. The three databases are called CNSIM1, CNSIM2, CNSIM3 and are available as three different resources: mySQL database, SPSS file and CSV file (see Figure \@ref(fig:testResources)). This example mimics real situations where different hospitals or research centers manage their own databases containing harmonized data. Data correspond to three simulated datasets with different numbers of observations of 11 harmonized variables. They contain synthetic data based on a model derived from the participants of the 1958 Birth Cohort, as part of an obesity methodological development project. This dataset does contain some NA values. The available variables are:


```{r insert_table_variables, echo=FALSE}
vars <- readr::read_delim("fig/table_variables_cnsim.txt", delim=",")
kable(vars)
```


The analyses that are described here, can also be found in the [DataSHIELD Tutorial](https://data2knowledge.atlassian.net/wiki/spaces/DSDEV/pages/714571780/Tutorial+for+DataSHIELD+users+v5+-+http+bit.ly+intro-DS) where these resources here uploaded to the Opal server as three tables, an inferior approach since data have to be moved from their original repositories.  

### Analysis from a single study

Let us start by illustrating how to analyze one data set (CNSIM2). 


```{r cnsim1}
library(DSOpal)
library(dsBaseClient)

# prepare login data and resource to assign
builder <- DSI::newDSLoginBuilder()
builder$append(server = "study1", url = "https://opal-demo.obiba.org", 
               user = "dsuser", password = "password", 
               resource = "RSRC.CNSIM1", driver = "OpalDriver")
logindata <- builder$build()

# login and assign resource
conns <- DSI::datashield.login(logins = logindata, assign = TRUE, 
                               symbol = "res")


# coerce ResourceClient objects to a data.frame called 'D'
datashield.assign.expr(conns, symbol = "D", 
                       expr = quote(as.resource.data.frame(res, strict=TRUE)))
```

Then we can inspect the type of data we have

```{r view_D}
ds.class("D")
ds.colnames("D")
```

Perform some data descriptive analyses

```{r descr_D}
ds.table("D$DIS_DIAB")
ds.table("D$DIS_DIAB", "D$GENDER")
```

Or even some statistical modelling. In this case we want to assess whether sex (GENDER) or triglycerides (LAB_TRIG) are risk factors for diabetes (DIS_DIAB)

```{r glm_D}
mod <- ds.glm(DIS_DIAB ~ LAB_TRIG + GENDER, data = "D" , family="binomial")
mod$coeff
```

As usual the connection must be closed

```{r close_conn_D}
datashield.logout(conns)
```


### Analysis from multiple studies

Now, let us illustrate a similar analysis with multiple studies. In this case we see results aggregated across all three studies.

```{r cnsim_multiple}
library(DSOpal)
library(dsBaseClient)

# prepare login data and resources to assign
builder <- DSI::newDSLoginBuilder()
builder$append(server = "study1", url = "https://opal-demo.obiba.org", 
               user = "dsuser", password = "password", 
               resource = "RSRC.CNSIM1", driver = "OpalDriver")
builder$append(server = "study2", url = "https://opal-demo.obiba.org", 
               user = "dsuser", password = "password", 
               resource = "RSRC.CNSIM2", driver = "OpalDriver")
builder$append(server = "study3", url = "https://opal-demo.obiba.org", 
               user = "dsuser", password = "password", 
               resource = "RSRC.CNSIM3", driver = "OpalDriver")
logindata <- builder$build()

# login and assign resources
conns <- datashield.login(logins = logindata, assign = TRUE, symbol = "res")

# assigned objects are of class ResourceClient (and others)
ds.class("res")

# coerce ResourceClient objects to data.frames
# (DataSHIELD config allows as.resource.data.frame() assignment function 
# for the purpose of the demo)
datashield.assign.expr(conns, symbol = "D", 
                       expr = quote(as.resource.data.frame(res, strict = TRUE)))
ds.class("D")

# do usual dsBase analysis
ds.summary('D$LAB_HDL')

# vector types are not necessarily the same depending on the data reader that was used
ds.class('D$GENDER')
ds.asFactor('D$GENDER', 'GENDER')
ds.summary('GENDER')

mod <- ds.glm("DIS_DIAB ~ LAB_TRIG + GENDER", data = "D" , family="binomial")
mod$coeff
```

*NOTE*: the outcome variable must be a 0/1 or a factor variable with two categories. The DataSHIELD function `ds.asFactor()` can be used to solve problems when having character variables. You can create a new object in the server that is a factor variable that can be passed in the left side of the formula indicated by `~ `.


We can also perform a meta-analysis approach

```{r}
mod.meta <- ds.glmSLMA("DIS_DIAB ~ LAB_TRIG + GENDER", 
                       dataName = "D" , family="binomial")
mod.meta

datashield.logout(conns)
```


and visualize the results using some R packages

```{r}
library(meta)
betas <- mod.meta$output.summary$input.beta.matrix.for.SLMA
ses <- mod.meta$output.summary$input.se.matrix.for.SLMA
betas
ses

res <- metagen(betas["LAB_TRIG",], ses["LAB_TRIG",])
forest(res, layout="JAMA")
```

## Practical exercise

>
*EXERCISE*: We have access to 3 datasets corresponding to simulated data from UKBiobank available through CINECA study. This data reproduces the exact associations found at UKBiobank. Next table shows the data dictionary of XX selected variables. The three datasets are accessed in this Opal server (https://opal-demo.obiba.org/) in a project called **GWAS** as three different resources (named **ega_phenotypes_1**, **ega_phenotypes_2** and **ega_phenotypes_3**). 
>
Start by installing the required R packages by typing 
>
```
library(devtools)
install.packages("DSI")
install_github("datashield/dsBaseClient")
```
>
Then, looad the three resources in R as `data.frame`'s using the functions available in the `DSI` library and answer the next questions using the functions available at `dsBaseClient` package.
>
* Check that your loaded objects are of class `data.frame`
* How many individuals have been diagnosed with diabetes by doctor (variable `diabetes_diagnosed_doctor`)?
* Obtain the same information stratified by sex (Hint: create a 2x2 table).
* Create an histogram of the variable `height` by combining information across the three different datasets (Hint: type `?ds.histogram` to see how to get this plot).
* Create a correlation plot between `bmi` and `weight` combining data from the three studies (Hint: ?ds.scatterPlot).
* Compute the correlation between `bmi`and `weight`.
* Fit a regression model between `cholesterol` and `weight`. 
* Fit a regression model between diabetes (variable `diabetes_diagnosed_doctor`) and colesterol level (variable `cholesterol`). Note: remember that outcome variable (e.g. diabetes) must be a factor variable. 
* Fit the same model adjusted by `bmi`. Is still cholesterol associated with diabetes?
* Is there any interaction between cholesterol and sex adjusted by bmi?

